Sensor Clusterization in D-optimal Design in Infinite Dimensional Linear Bayesian Inverse Problems

### 2020-10-14 new_paper.tex
looking better and better

Designing experiments is usually done by optimizing some \emph{design
  criterion} and this is true both for frequentists
\cite{Silvey13,Ucinski05} as well as for Bayesians
\cite{ChalonerVerdinelli95}, see \cite{ChalonerVerdinelli95} for an
investigation of the analogy between the two approaches. Although
there is a plethora of design criteria, in the current work we will
focus on just one of these, commonly referred to as \emph{D-optimal
  design}. It has a simple and appealing motivation in the Bayesian
context as explained in \cite{ChalonerVerdinelli95}: for a linear
model, a D-optimal design seeks to maximize the expected information
gain (KL divergence \cite{KullbackLeibler51,CoverThomas12}) between
posterior and prior. In finite dimensions, maximizing the expected
information gain amounts to minimizing the determinant of the
posterior covariance matrix. In a frequentist setting, an optimal
design minimizes the volume of the uncertainty ellipsoid \cite[page
  16]{Ucinski05}, but this is done for the Fisher information matrix
and not the posterior covariance. However, \cite{ChalonerVerdinelli95}
shows that the latter is just a regularized version of the former.

The previous discussion is classical for experimental design when
inference is to take place over a finite (not too large) number of
parameters. The subject of optimal experimental design for function
inference in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14,AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, it can be shown that a D-optimal design arises naturally for
linear models when one wishes to maximize the KL divergence between
posterior and prior and this amounts to minimizing the determinant of
the posterior covariance operator (understood as a product of its
eigenvalues). Some difficulties arise in the process, but remedies can
be found as shown in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D-optimal criterion (and others that will not be
investigated here), the optimization process results in measurements
that are very similar. For example, if a measurement is thought of as
measuring some function value at $\x \in \Omega \subseteq \R^d,
d=1,2,3$ (with added error) then the optimization procedure sometimes
places sensors in close proximity to each other (as can be seen
in figure \ref{fig:clusterization illustration}). Following
\cite{Ucinski05}, we refer to this phenomenon as \emph{sensor
  clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{Fedorov96} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design,
where the user enforces measurement locations to be far from each
other. One way to do this is by introducing correlated errors which,
philosophically, accounts for both measurement error and model error.
Another method considered is imposing distance constraints between
measurements. A somewhat different approach is suggested in \cite[page
  49]{Fedorov12}, where close by design measurements are merged --- a
procedure which obviously does not avoid clusterization. The same problem
arises in time-series analysis for pharmacokinetic experiments. The
authors of \cite{Hooker09} suggest modeling auto-correlation time in
the noise model, which is equivalent to the correlated errors
mentioned above.

@@

\subsection{Contribution}
We propose and thoroughly study an analytically tractable model for
understanding D-optimal designs. In this model, an optimal design
arises as a solution to a constrained optimization problem, formulated
using Lagrange multipliers (section \ref{section:D and grad}). In the
case of no model error we show this problem is in fact a set of
eigenvector-eigenvalue problems (equation \eqref{eq:eigenproblem}). We
then characterize an optimal design as one that reduces the highest
uncertainties, and show how that can give rise to repeated
measurements (section \ref{section:clusterization} and figure
\ref{fig:clusterization}).

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings --- the Matrix Determinant Lemma and a
lemma for calculating the derivative of determinants. We also provide
other tools for understanding D-optimal designs in
infinite-dimensional Bayesian inverse problems, including lemmas for
calculating the increase in the design criterion per observation ---
Lemmas \ref{lemma:design increase} and Corollary \ref{cor:zero mod
  err}. %%  We also show that in our model with independent measurement
%% noise, an interesting phenomenon arises, in which the eigenvectors
%% of the measurement operator (denoted below $\obs^*\obs$) are the
%% same as the eigenvectors of the prior covariance in observation
%% space, which helps arrive at the results outlined in the previous
%% paragraph.

\subsection{Limitations}\label{subsec:limitations}
There are two main drawbacks of the arguments presented here. First,
one can argue that we allow considerably less restrictive observations
than any real-life problem does. This is because our observation model
allows taking measurements that are arbitrarily weighted linear
combinations of eigenvectors. The second drawback is that we do not
show rigorously that clusterization necessarily occurs in our model,
we only show that there is no reason to be surprised when it does
occur.

@@

@@

\begin{restatable}[Characterizing D-optimal designs]{theorem}{char}\label{thm:char}
  Let:
  \begin{enumerate}
  \item $\fwd,\prcov,\obs$ the forward, prior covariance and observation operators, respectively.
  \item $\{\lambda_i\}_{i=1}^\infty$ the eigenvalues of
    $\fwd\prcov\fwd^*$ in decreasing order.
    \item $\{\ev_i\}_{i=1}^\infty$ the corresponding eigenvectors.
  \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$, where
    $\obs$ maximizes $\tar(\obs)$.
  \item $m$ the number of measurements taken.
  \item $\sigma^2$ observation noise variance.
  \end{enumerate}
  Then:
  \begin{enumerate}
  \item $\obs^*\obs$ and $\fwd\prcov\fwd^*$ are simultaneously diagonalizable.
  \item There exists $0 < k \leq m$ such that $\eta_i = 0$ for $i\geq k+1$.
  \item $\sigma^2\lambda_i^{-1} + \eta_i$ are equal for all $1 \leq i \leq k$.
  \item If $k < m$, then $\sigma^2\lambda_i^{-1} + \eta_i \leq
    \sigma^2\lambda_{k+1}^{-1}, i=1,\dots,k$.
  \end{enumerate}
\end{restatable}

\section{Sensor Clusterization --- Vanishing Model Error}\label{section:clusterization}
\subsection{Overview}
In this section we use theorem \ref{thm:char} to explain how sensor
clusterization may arise in various scenarios. We first consider a
sequential optimal design. We then consider the case where all
measurements are chosen at once. Finally, we hypothesize how this may
apply to a real-life inverse problem.

\subsection{Sequential Design}
Recall the sequential optimal design scheme from section
\ref{subsec:seq vs sim} and the definition of $\obs_k$ from
\eqref{eq:def obs_k}. The way clusterization can occur in this
scenario is understood using thorem \ref{thm:char}, specialized to one
measurement each time. In this case $m=1$, so $k=1$, so $\meas_1
\parallel \ev_1$. As hinted in section \ref{subsec:seq vs sim}, the
posterior becomes the prior for the next step. Similar reasoning leads
to:
\begin{align*}
  \begin{split}
    \meas_2 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_2 \\
    \meas_2 \parallel \ev_2  &\text{ o.w. }
  \end{split}
\end{align*}
In the first case, we get clusterization. In the second, we consider $\meas_3$:
\begin{align*}
  \begin{split}
    %\meas_3 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_3 \text{ (note $\lambda^{-1}_1 < \lambda^{-1}_2$)}  \\
    %\meas_3 \parallel  \ev_2  &\text{ if } \\
    \meas_3 \parallel  \ev_3  &\text{ o.w. }
  \end{split}
\end{align*}
Again, the first case results in clusterization. The sequential design
proceeds in this fashion, where every $\meas_i$ is taken in the
direction of the eigenvector of smallest eigenvalue of the previous
posterior precision operator. Taking enough measurements we will
surely arrive at clusterization, since the prior precision increases
quickly --- in section \ref{subsec:abstract OED} we assumed  $\fwd$
is strongly smoothing, which typpically means eigenvalues of $(\fwd
\prcov\fwd^*)^{-1}$ grow exponentially.

\subsection{Simultaneous Design}
In a simultaneous design scenario, the result is not as clear cut as
in the sequential design case. We still rely on previous results ---
namely corollary \ref{cor:same ev} and theorem \ref{thm:char}. An
illustration of two D-optimal designs scenarios for the same problem
are presented in figure \ref{fig:clusterization} --- the first
exhibits clusterization and second does not. As was discussed earlier
and illustrated in figure \ref{fig:optimal vs not}, an optimal design
makes the precision ``flat'' where it is lowest. Clusterization may
arise when two or more eigenvectors have equal precision. This
happens, for example for eigenvectors $1$ and $2$ when considering
measurements $\meas_i,i=1,2,3$. Taking three measurements, it is
possible that $\meas_1$ is such that it equalizes the precision of
eigenvectors $1$ and $2$. In this case, taking mesurements $\meas_2 =
\meas_3 = \frac{1}{2}(\ev_1 + \ev_2)$ leads to an optimal design that
exhibits clusterization. Similarly, taking $\meas_5 = \meas_6 =
\frac{1}{3}(\ev_1 + \ev_2 + \ev_3)$ also gives rise to an optimal
design that exhibits clusterization. It is important to note that
there are (infinitely) many optimal designs that actually do not
exhibit clusterization --- e.g. the second design presented in figure
\ref{fig:clusterization}. In the following section we postulate what
causes such designs to be observed in realistic experimental design
scenarios.

@@

\subsection{Why do Realistic Designs Exhibit Clusterization?}
It is unclear from the previous analysis what makes realistic designs
exhibit clusterization, such as the one observed in figure
\ref{fig:clusterization illustration}. We can only hypothesize.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Non-Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis of Optimal Designs --- Non-Vanishing Model Error}\label{section:non vanishing}
\subsection{Overview}
In this section we show the effect model error has on the
clusterization phenomenon. In contrast to the previous case of
vanishing model error, we will see that if $\modcov \neq 0$ clustering
will not occur.

\subsection{Derivation}
We rely on corollary \ref{cor:same meas} below, which is proved in the
appendix. Denote $\obs = (\meas_1,\dots,\meas_m)^t$ and $\obsm :=
(\meas_1,\dots,\meas_{m-1})^t$.
\begin{restatable*}{corollary}{samemeas}\label{cor:same meas}
  If $\meas_m = \meas_j$ for some $1 \leq j \leq m-1$, then
  \begin{equation*}
    \tar(\obs) - \tar(\obsm) =
    \log \left ( 1 + \frac{\sigma^2
      \langle \fwd \postcovm \fwd^* \obsm^* \Sigmam^{-1} e_j,
      \obsm^* \Sigmam^{-1}e_j \rangle
    }{
      2 - \sigma^2 e_j^t\Sigmam^{-1}e_j 
    }       
    \right ).
  \end{equation*}
\end{restatable*}
Recall from \eqref{eq:Sigma} that $\Sigma(\obs) = \obs \modcov \obs^*
+ \sigma^2I$. From Corollary \ref{cor:same meas} we can deduce that

@@

Hence no increase in the design criterion is achieved by taking a
repeated measurement. This is clearly sub-optimal and $\obsm$ cannot
be an optimal design. Thus, for small measurement error levels, the
clusterization effect will be mitigated by the presence of a non-zero
model error. Since the design criterion is not defined for $\sigma^2 =
0$ and identical measurements, we cannot make a statement regarding
$\sigma^2 = 0$, except in the limiting sense described above.

@@

\section{Acknowledgements}
This study is a part of my PhD thesis \cite{mine} which was written
under the instruction of Prof. Georg Stadler in New York University's
Courant Institute. I would like to thank him for his great mentorship.

@@

\begin{lemma}[Matrix Determinant Lemma in Hilbert Spaces]\label{lemma:MDL}
  Let $\hilp$ a separable Hilbert space, $u,v\in \hilp$ and $A: \hilp \to
  \hilp$ an invertible linear operator such that $\tr{A-I} <
  \infty$. Then $\det A$ and $\det A + uv^*$ are well defined and
  \begin{equation*}
    \det (A + uv^*) = (1 + \langle A^{-1} u, v \rangle ) \det A,
  \end{equation*}
  where $(A + uv^*)w := Aw + \langle v,w \rangle u$.
\end{lemma}
\begin{proof}
  In this proof we rely on definitions and results from
  \cite{Simon77}. First, consider $B := I + xy^*$ for some $x,y \in
  \hilp$. We construct an eigenbasis for $B$ and use that to show $\det
  B = 1 + \langle x, y \rangle$. First let $x_1 := x$.  Now, if $x
  \parallel y$, take $\{x_n \}_{n=2}^{\infty}$ an orthogonal basis for
  $span\{x_1\} ^{\perp}$. If, on the other hand, $x \nparallel y$, let
  \begin{equation*}
    x_2 := x - \frac{ \langle x, y\rangle}{\|y\|^2}y
  \end{equation*}
  and it is easy to verify that $x_2 \perp y$ and $span \{x,y\} = span
  \{x_1,x_2\}$. Take $\{x_n \}_{n=3}^{\infty}$ an orthogonal basis for
  $span\{x_1,x_2\} ^{\perp}$. In both cases,
  \begin{equation*}
    B x_n =
    \begin{cases}
      (1 + \langle x, y \rangle) x_n & n = 1 \\
      x_n                            & n \neq 1,
    \end{cases}
  \end{equation*}
  and so $\det B = 1 + \langle x, y \rangle$.

@@

  Define $C = S^tS - I \in \R^{m \times m}$. Note that $\ttr C = 0$ and
  $C$ is diagonal with non-zero entries $d_i-1,i=1,\dots,k$. It suffices
  to find $V$ orthogonal such that $V C V^t$ has zero diagonal. We
  construct such $V$ by sequentially inserting zeros in the diagonal
  and not destroying zeros we already introduced, starting from the
  last diagonal entry and moving to the first. If $c_{mm} \neq 0$ ,
  let $p < m$ such that $c_{pp}c_{mm} < 0$ (such $p$ exists because
  the trace is zero) and let $\theta \in (0,\pi)$. Define a Givens
  rotation $R^{(m)} \in \R^{m \times m}$ by
  \begin{equation*}
    r^{(m)}_{ab} :=
    \begin{cases}
      1 & a = b \neq p \text{ or } a = b \neq m \\
      \cos \theta & a = b = p  \\
     -\sin \theta & a = p, b = m\\
      \cos \theta & a = b = m \\
      \sin \theta & a = m, b = p \\ 
      0 & o.w
    \end{cases}
  \end{equation*}
  Note that conjugating a matrix by $R^{(m)}$ changes only its $m$ and
  $p$ rows and columns. We want to choose $\theta$ such that
  \begin{equation}\label{eq:mm}
    0 = [R^{(m)} C (R^{(m)})^t]_{mm} = \cos^2 \theta c_{mm} + 2\cos \theta \sin
    \theta c_{mp} + \sin^2\theta c_{pp},
  \end{equation}
  and it suffices to choose $\theta$ such that
  \begin{equation*}
    c_{mm} \cot^2 \theta + 2 c_{mp} \cot \theta + c_{pp} = 0.
  \end{equation*}
  This quadratic in $\cot\theta$ has a real solution, since
  $c_{pp}c_{mm} < 0$ by assumption and we can find $\theta \in
  (0,\pi)$ such that \eqref{eq:mm} is satisfied. We continue to find
  $R^{(m-1)}$ that leaves row and column $m$ unchanged and
  continue introducing zeros to the diagonal. The assumption $\ttr D =
  m \Rightarrow \ttr C = 0$ guarantees we can do that. Taking $V:=
  R^{(1)} R^{(2)} \dots R^{(m-1)}R^{(m)}$ completes the proof.
\end{proof}

@@

%    Bibliographies can be prepared with BibTeX using amsplain,
%    amsalpha, or (for "historical" overviews) natbib style.
\bibliographystyle{amsplain}
\bibliography{refs.bib,georg_refs}

>>>>

Designing experiments is usually done by optimizing some \emph{design
  criterion}. This is true both for frequentists
\cite{Silvey13,Ucinski05} as well as for Bayesians
\cite{ChalonerVerdinelli95}. See \cite{ChalonerVerdinelli95} for an
investigation of the analogy between the two approaches. Although
there is a plethora of design criteria, we focus on just one of these,
commonly referred to as \emph{D-optimal design}. It has a simple and
appealing motivation in the Bayesian context as explained in
\cite{ChalonerVerdinelli95}: for a linear model, a D-optimal design
seeks to maximize the expected information gain (KL divergence
\cite{KullbackLeibler51,CoverThomas12}) between posterior and
prior. In finite dimensions, maximizing the expected information gain
amounts to minimizing the determinant of the posterior covariance
matrix. In a frequentist setting, a D-optimal design minimizes the
volume of the uncertainty ellipsoid \cite[page 16]{Ucinski05}, but
this is done for the Fisher information matrix and not the posterior
covariance. However, \cite{ChalonerVerdinelli95} show that the latter
is just a regularized version of the former.

The previous discussion is classical for experimental design when
inference is to take place over a finite (not too large) number of
parameters. The subject of optimal experimental design for function
inference in a Bayesian context was pioneered by
\cite{AlexanderianGloorGhattas14, AlexanderianPetraStadlerEtAl16,
  AlexanderianPetraStadlerEtAl14}. Similarly to the finite dimensional
case, it can be shown that a D-optimal design arises naturally for
linear models when one wishes to maximize the KL divergence between
posterior and prior. This amounts to minimizing the determinant of the
posterior covariance operator (understood as a product of its
eigenvalues). Some difficulties arise in the process, but remedies can
be found as shown in \cite{AlexanderianGloorGhattas14}.

It seems counter intuitive that when one computes an optimal design
using the D-optimal criterion (and others that will not be
investigated here), the optimization process results in measurements
that are very similar. For example, if a measurement is thought of as
measuring some function value at $\x \in \Omega \subseteq \R^d,
d=1,2,3$ (with added error) then the optimization procedure sometimes
places sensors in very close proximity to each other (as can be seen
in figure \ref{fig:clusterization illustration}). Following
\cite{Ucinski05}, we refer to this phenomenon as \emph{sensor
  clusterization}.

\subsection{Related Work}
The phenomenon of sensor clusterization seems to be known in several
different contexts. In a frequentist and finite-dimensional context,
\cite{Fedorov96} and \cite[chapter 2.4.3]{Ucinski05} discuss this
phenomenon and suggest an approach called clusterization-free design.
In such designs, the user enforces measurement locations to be far
from each other. One way to do this is by introducing correlated
errors which, philosophically, accounts for both measurement error and
model error. Another method considered is imposing distance
constraints between measurements. A somewhat different approach is
suggested in \cite[page 49]{Fedorov12}, where close by design
measurements are merged --- a procedure which obviously does not avoid
clusterization. The same problem arises in time-series analysis for
pharmacokinetic experiments. The authors of \cite{Hooker09} suggest
modeling auto-correlation time in the noise model, which is equivalent
to the correlated errors mentioned above.

@@

\subsection{Contribution}
We propose and thoroughly study an analytically tractable model for
understanding D-optimal designs. In this model, an optimal design
arises as a solution to a constrained optimization problem, formulated
using Lagrange multipliers (section \ref{section:D and grad}). Using
this model we rigorously show how model error mitigates clusterization
(section \ref{section:non vanishing}). We show how clusterization
occurs in a sequential design scenario (section
\ref{subsec:clusterization sequential}). Then, we show that designs
that exhibit clusterization are just as optimal as clusterization free
designs in a simultaneous D-optimal designs (section
\ref{subsec:clusterization simultaneous}).

A beautiful mathematical structure arises in D-optimal designs when no
model error is present. The Lagrange multipliers problems is in fact a
``weird'' eigenvalue problem (``weird'' because the operator depends
on its eigenvectors, see \eqref{eq:eigenproblem}). This helps us shed
light into D-optimal designs. We characterize these designs using
theorem \ref{thm:char}. One key and intuituve insight we prove is that
a D-optimal design reduces uncertainties where they are highest
first. More interestig phenomena arise, but they require setting
notation, so these are discussed later.

In the process we generalize several lemmas from linear algebra to
infinite-dimensional settings. A matrix determinant emma and a lemma
due to Lax \cite{lax97} for calculating the derivative of determinants
of certain operators (\ref{lemma:MDL} and \ref{lemma:lax},
respectfully). We also prove some non-trivial lemmas in linear
algebra. One of which (\ref{lemma:free}) gives a decomposition $M =
AA^t$ where $A$ has unit norm columns. Another (\ref{lemma:sim diag})
shows simultaneous diagonizability for the ``weird'' eigenvalue problem
referenced above. We also provide other tools for understanding
D-optimal designs in infinite-dimensional Bayesian inverse problems,
including lemmas for calculating the increase in the design criterion
per observation --- Lemmas \ref{lemma:design increase} and Corollary
\ref{cor:zero mod err}.

\subsection{Limitations}\label{subsec:limitations}
There are two main drawbacks of the study presented here. Our relaxed
model does not consider any specific set of allowed
measurements. Rather, we take measurements in the unit ball in some
Hilbert space. This allows considerably less restrictive observations
than any real-life problem does. The second drawback is that we do not
show rigorously that clusterization necessarily occurs in a
simultaneous design. We only show that it is as reasonable as no
clusterization.

@@

\begin{theorem}[Characterizing D-optimal designs]\label{thm:char}
  Let:
  \begin{enumerate}
  \item $\fwd, \prcov, \obs$ the forward, prior covariance and
    observation operators, respectively.
  \item $\{\lambda_i\}_{i=1}^\infty$ the eigenvalues of
    $\fwd\prcov\fwd^*$ in decreasing order.
  \item $\{\ev_i\}_{i=1}^\infty$ the corresponding eigenvectors.
  \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$, where
    $\obs$ maximizes $\tar(\obs)$.
  \item $m$ the number of measurements taken.
  \item $\sigma^2$ observation noise variance.
  \end{enumerate}
  Then:
  \begin{enumerate}
  \item $\obs^*\obs$ and $\fwd\prcov\fwd^*$ are simultaneously diagonalizable.
  \item There exists $0 < k \leq m$ such that $\eta_i = 0$ for $i\geq k+1$.
  \item $\sigma^2\lambda_i^{-1} + \eta_i$ are equal for all $1 \leq i \leq k$.
  \item If $k < m$, then $\sigma^2\lambda_i^{-1} + \eta_i \leq
    \sigma^2\lambda_{k+1}^{-1}, i=1,\dots,k$.
  \end{enumerate}
\end{theorem}

@@

\section{Sensor Clusterization --- Vanishing Model Error}\label{section:clusterization}
\subsection{Overview}
In this section we use theorem \ref{thm:char} to explain how sensor
clusterization may arise in various scenarios. We first consider a
sequential optimal design. Then we consider simultaneous
designs.

\subsection{Sequential Design}\label{subsec:clusterization sequential}
Recall the sequential optimal design scheme from section
\ref{subsec:seq vs sim} and the definition of $\obs_k$ from
\eqref{eq:def obs_k}. The way clusterization can occur in sequential
design scenario is understood by specializing thorem \ref{thm:char} to
one measurement each time. In this case $m=1$, so $k=1$, $\meas_1
\parallel \ev_1$ and $\eta_1 = 1$. As hinted in section
\ref{subsec:seq vs sim}, the posterior becomes the prior for the next
step. Similar reasoning leads to:
\begin{align*}
  \begin{split}
    \meas_2 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_2 \\
    \meas_2 \parallel \ev_2  &\text{ o.w. }
  \end{split}
\end{align*}
In the first case, we get clusterization. In the second, we consider $\meas_3$:
\begin{align*}
  \begin{split}
    \meas_3 \parallel \ev_1  &\text{ if } \sigma^2 \lambda^{-1}_1 + 1 < \sigma^2\lambda^{-1}_3 \\
    %\meas_3 \parallel  \ev_2  &\text{ if } \\
    \meas_3 \parallel  \ev_3  &\text{ o.w. }
  \end{split}
\end{align*}
Again, the first case results in clusterization. The sequential design
proceeds in this fashion. Every $\meas_i$ is taken in the direction of
the eigenvector of smallest eigenvalue of the previous posterior
precision operator. Taking enough measurements we will surely arrive
at clusterization, since the prior precision increases quickly --- in
section \ref{subsec:abstract OED} we assumed $\fwd$ is strongly
smoothing, which means eigenvalues of $(\fwd \prcov\fwd^*)^{-1}$ grow
quickly.

\subsection{Simultaneous Design}\label{subsec:clusterization simultaneous}
In a simultaneous design scenario, the result is not as clear cut as
in the sequential design case. We still rely on previous results ---
namely corollary \ref{cor:same ev} and theorem \ref{thm:char}. An
illustration of two D-optimal designs scenarios for the same problem
are presented in figure \ref{fig:clusterization} --- the first
exhibits clusterization and second does not. As was discussed earlier
and illustrated in figure \ref{fig:optimal vs not}, an optimal design
makes the precision ``flat'' where it is lowest. Clusterization may
arise when two or more eigenvectors have equal precision. This
happens, for example for eigenvectors $1$ and $2$ when considering
measurements $\meas_i,i=1,2,3$. Taking three measurements, it is
possible that $\meas_1$ is such that it equalizes the precision of
eigenvectors $1$ and $2$. In this case, taking mesurements $\meas_2 =
\meas_3 = \frac{1}{2}(\ev_1 + \ev_2)$ leads to an optimal design that
exhibits clusterization. Similarly, taking $\meas_5 = \meas_6 =
\frac{1}{3}(\ev_1 + \ev_2 + \ev_3)$ also gives rise to an optimal
design that exhibits clusterization. It is important to note that
there are (infinitely) many optimal designs that actually do not
exhibit clusterization --- e.g. the second design presented in figure
\ref{fig:clusterization}. In the following section we postulate what
causes such designs to be observed in realistic experimental design
scenarios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% SECTION Analysis of Optimal Designs --- Non-Vanishing Model Error
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis of Optimal Designs --- Non-Vanishing Model Error}\label{section:non vanishing}
In this section we show the effect model error has on the
clusterization phenomenon. We will see that if $\modcov \neq 0$
clustering will not occur. This is contrary to the previous case of
vanishing model error.

Corollary \ref{cor:same meas} below is proved in the appendix. Denote
$\obs = (\meas_1,\dots,\meas_m)^t$ and $\obsm :=
(\meas_1,\dots,\meas_{m-1})^t$. Every symbol with a wide hat
$\ \widehat{}\ $ refers to the marked object for $\obsm$. E.g. $\Sigmam =
\Sigma(\obsm)$.
\begin{restatable*}{corollary}{samemeas}\label{cor:same meas}
  If $\meas_m = \meas_j$ for some $1 \leq j \leq m-1$, then
  \begin{equation*}
    \tar(\obs) - \tar(\obsm) =
    \log \left ( 1 + \frac{\sigma^2
      \langle \fwd \postcovm \fwd^* \obsm^* \Sigmam^{-1} e_j,
      \obsm^* \Sigmam^{-1}e_j \rangle
    }{
      2 - \sigma^2 e_j^t\Sigmam^{-1}e_j 
    }       
    \right ).
  \end{equation*}
\end{restatable*}
Recall from \eqref{eq:Sigma} that $\Sigma(\obs) = \obs \modcov \obs^*
+ \sigma^2I$. Then

@@

Hence no increase in the design criterion is achieved by taking a
repeated measurement (in the limit $\sigma^2 \to 0$). This is clearly
sub-optimal and $\obsm$ cannot be an optimal design. Thus, for small
measurement error levels, the clusterization effect will be mitigated
by the presence of a non-zero model error. Since the design criterion
is not defined for $\sigma^2 = 0$ and identical measurements, we
cannot make a statement regarding $\sigma^2 = 0$, except in the
limiting sense described above.

@@

\section{Acknowledgements}
This study is a part of my PhD thesis. It was written under the
instruction of Prof. Georg Stadler in New York University's Courant
Institute. I would like to thank him for his great mentorship.

@@

\begin{lemma}[Matrix Determinant Lemma in Hilbert Spaces]\label{lemma:MDL}
  Let $\hil$ a separable Hilbert space, $u,v\in \hil$ and $A: \hil \to
  \hil$ an invertible linear operator such that $\tr{A-I} <
  \infty$. Then $\det A$ and $\det A + uv^*$ are well defined and
  \begin{equation*}
    \det (A + uv^*) = (1 + \langle A^{-1} u, v \rangle ) \det A,
  \end{equation*}
  where $(A + uv^*)w := Aw + \langle v,w \rangle u$.
\end{lemma}
\begin{proof}
  In this proof we rely on definitions and results from
  \cite{Simon77}. First, consider $B := I + xy^*$ for some $x,y \in
  \hil$. We construct an eigenbasis for $B$ and use that to show $\det
  B = 1 + \langle x, y \rangle$. First let $x_1 := x$.  Now, if $x
  \parallel y$, take $\{x_n \}_{n=2}^{\infty}$ an orthogonal basis for
  $span\{x_1\} ^{\perp}$. If, on the other hand, $x \nparallel y$, let
  \begin{equation*}
    x_2 := x - \frac{ \langle x, y\rangle}{\|y\|^2}y
  \end{equation*}
  and it is easy to verify that $x_2 \perp y$ and $span \{x,y\} = span
  \{x_1,x_2\}$. Take $\{x_n \}_{n=3}^{\infty}$ an orthogonal basis for
  $span\{x_1,x_2\} ^{\perp}$. In both cases,
  \begin{equation*}
    B x_n =
    \begin{cases}
      (1 + \langle x, y \rangle) x_n & n = 1 \\
      x_n                            & n \neq 1,
    \end{cases}
  \end{equation*}
  and so $\det B = 1 + \langle x, y \rangle$.

@@

  Define $C = S^tS - I \in \R^{m \times m}$. Note that $\ttr C = 0$ and
  $C$ is diagonal with non-zero entries $d_i-1,i=1,\dots,k$. It suffices
  to find $V$ orthogonal such that $V C V^t$ has zero diagonal. We
  construct such $V$ by sequentially inserting zeros in the diagonal
  and not destroying zeros we already introduced, starting from the
  last diagonal entry and moving to the first. Since $c_{mm} \neq 0$ ,
  let $p < m$ such that $c_{pp}c_{mm} < 0$ (such $p$ exists because
  the trace is zero) and let $\theta \in (0,\pi)$. Define a Givens
  rotation $R^{(m)} \in \R^{m \times m}$ by
  \begin{equation*}
    r^{(m)}_{ab} :=
    \begin{cases}
      1 & a = b \neq p \text{ or } a = b \neq m \\
      \cos \theta & a = b = p  \\
     -\sin \theta & a = p, b = m\\
      \cos \theta & a = b = m \\
      \sin \theta & a = m, b = p \\ 
      0 & o.w
    \end{cases}
  \end{equation*}
  Note that conjugating a matrix by $R^{(m)}$ changes only its $m$ and
  $p$ rows and columns. We want to choose $\theta$ such that
  \begin{equation}\label{eq:mm}
    0 = [R^{(m)} C (R^{(m)})^t]_{mm} = \cos^2 \theta c_{mm} + 2\cos \theta \sin
    \theta c_{mp} + \sin^2\theta c_{pp},
  \end{equation}
  and it suffices to choose $\theta$ such that
  \begin{equation*}
    c_{mm} \cot^2 \theta + 2 c_{mp} \cot \theta + c_{pp} = 0.
  \end{equation*}
  This quadratic in $\cot\theta$ has a real solution, since
  $c_{pp}c_{mm} < 0$ by assumption and we can find $\theta \in
  (0,\pi)$ such that \eqref{eq:mm} is satisfied. We continue to find
  $R^{(m-1)}$ that leaves row and column $m$ unchanged and
  continue introducing zeros to the diagonal. The assumption $\ttr D =
  m \Rightarrow \ttr C = 0$ guarantees we can do that. Taking $V:=
  R^{(1)} R^{(2)} \dots R^{(m-1)}R^{(m)}$ completes the proof.
\end{proof}

@@

%    Bibliographies can be prepared with BibTeX using amsplain,
%    amsalpha, or (for "historical" overviews) natbib style.
\bibliographystyle{amsplain}
\bibliography{refs,georg_refs}
